---
title: "ESM 244 Lab 1"
author: "Stephanie Copeland"
date: "1/6/2022"
output: html_document
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
#^takes care of this chunk
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
#^takes care of all chunks

#attach packages
library(tidyverse)
library(here)
library(sf)
library(tmap)
### install.packages('tmap')
### ^can highlight just code chunk behind #sign to run
### update.packages(ask = FALSE)
```

cmd-shift-enter = shortcut for running the current code chunk

### Read in data

cmd-option-i = short cut for making code chunk

```{r}
sf_trees <- read_csv(here('data', 'sf_trees', 'sf_trees.csv'))
#show_col_types = FALSE, if you want thing below to be quiet
```

### Part 1: Wrangling and ggplot review

**Example 1** Find counts of observations by `legal_status` & wrangle a bit

```{r}
### method 1: group_by() %>% summarize()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

### method 2: different way plus a few new functions
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% #dropping NAs in legal status column 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>%  #brings to front
  slice_max(tree_count, n = 5) %>%  #takes top 5 but could be any number
  arrange(-tree_count) #highest to lowest or desc(tree_count)
```

Make a graph of the top 5 from above
```{r}
# put - reverses fct_reorder arrangement 
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) +
  geom_col(fill = 'darkgreen') +
  labs(x = 'Legal status', y = 'Tree count') +
  coord_flip() +
  theme_minimal()
```

**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretakes is "MTA", and store as `permitted_data_df`

shift-cmd-c to comment/uncomment quickly

```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker)
permitted_data_df <- sf_trees %>% 
  filter(legal_status == 'Permitted Site', caretaker == 'MTA')
# an & instead of , would do the same thing
# | is the "or" instead of the and function so permitted site OR caretake too
```

**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `data`, `latitude`, `longitude`, and store as `blackwood_acadia_df`

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)

### make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = 'darkgreen')
  
```

**Example 4:** use `tidyr::separate()`

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')
```

**Example 5:** use `tidyr::unite`
```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = '_COOL_')
```

